Tight Bounds for Dynamic Balanced Graph Partitioning

### 2020-05-12 paper/disc20clustering.tex
tildes

\begin{abstract}
	Distributed   applications,  including  batch  processing, streaming, scale-out databases,
	or machine learning, generate a significant amount of network traffic.
	By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), we can reduce the network load and  improve application performance. 
	However, the communication pattern of different applications is often unknown a priori and may change over time, hence it needs to be learned in an online manner.
	%
	This paper revisits the online 
	balanced partitioning problem 
	(introduced by Avin et al.~at DISC 2016)
	that asks for an algorithm that strikes
	an optimal tradeoff between the benefits
	of collocation (i.e., lower network load) 
	and its costs (i.e., migrations). 
	%
	Our first contribution is a significantly improved
	lower bound of $\Omega(k\cdot \ell)$ on the
	competitive ratio, where $\ell$ is the number
	of clusters and $k$ is the cluster size,
	even for a scenario in which the communication
	pattern is static and can be perfectly partitioned;
	we also provide an asymptotically tight upper bound 
	of $O(k\cdot \ell)$ for this scenario.
	For $k=3$, we contribute an asymptotically tight upper bound
	of $\Theta(\ell)$
	for the general model in which the
	communication pattern can change arbitrarily over time.
	We improve the result for $k=2$ by providing a strictly $6$-competitive upper bound for the general model.
	In contrast to most prior work, our algorithms respect all capacity constraints and do not require resource augmentation.

@@

The popularity of data-centric, distributed applications has led to an explosive growth of network traffic, especially in data centers \cite{roy2015inside,singh2015jupiter}.
The performance of these distributed applications often critically depends on the underlying network \cite{mogul2012we}, and efficient operation of these networks is important.
At the same time, distributed systems are often highly virtualized today, and provide interesting new opportunities for resource optimization.
In particular, it has become possible to operate data centers in a more demand-aware manner: 
by dynamically migrating nodes (e.g., virtual machines) which communicate frequently topologically closer to each other, network traffic can be reduced significantly.  
However, migrations should be used moderately, as migrations also entail overheads.

This paper studies the algorithmic problem underlying such demand-aware
optimizations, aiming to strike a balance between the benefits of migrations (e.g., reduced network load) and their costs.
In particular, we are interested in an online variant of the problem: since communication patterns can change over time, an online algorithm needs to react dynamically to new traffic patterns, and migrate nodes  accordingly.
Ideally, this algorithm should perform close to an optimal offline algorithm, without requiring any information about future traffic demands.

This problem is known as the dynamic balanced graph partitioning problem and was introduced by Avin et al. \cite{repartition-disc, sidma-arxiv} at DISC 2016. A special variant of the general problem has later been studied by Henzinger et al. \cite{sigmetrics19_partitioning} at SIGMETRICS 2019.
We refer to the latter as the learning model.

@@

We study two models in this paper: the \emph{general partitioning} model, and its subproblem, the \emph{learning} model.
In both models, we assume that communication patterns are not known to our algorithms at the beginning.
We measure the~quality of~presented algorithmic solutions by competitive analysis~\cite{borodin-book}, which is well-suited for problems that are online by their nature.
In the competitive analysis, the goal is to~optimize \emph{the competitive ratio} of a given online algorithm: the ratio of its cost to the cost of~an~optimal offline algorithm that knows the whole input sequence in advance.
%We emphasize that all our algorithms have a \emph{strict} competitive ratio (without an additive term).

\noindent
\textbf{General partitioning model.}
In the \emph{dynamic balanced graph partitioning} problem, we are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a~sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

An algorithm serves a communication request between two nodes
either \emph{locally} at cost~0
if they are collocated,
or \emph{remotely} at cost~1
if they are located in different clusters.
We refer to these two types of requests as \emph{internal}
and \emph{external} requests, respectively.
Before serving a request,
an online algorithm may perform a \emph{repartition},
%(i.e., \emph{reconfigure}).
i.e.,
it may move (``migrate'') some nodes into clusters different from their current clusters, while respecting the capacity of every cluster. 
Afterward, 
the algorithm serves the  request.
The cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $\ALG$,
its cost,
denoted by $\ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $\ALG$ while serving the sequence $\sigma$.

\noindent
\textbf{Learning model.}
We study a \emph{learning} variant of dynamic balanced graph partitioning,
where the communication pattern is \emph{static}:
whether a pair of  nodes ever communicate or not, 
is determined a priori and is unknown to algorithms,
 and requests to such pairs arrive indefinitely.
Any algorithm must eventually collocate  pairs of communicating nodes,
as otherwise it cannot be competitive.
As in Henzinger et al. \cite{sigmetrics19_partitioning}, we assume that the communication graph admits a \emph{perfect partition},
i.e., a partition in which no inter-cluster request ever occurs.
%any algorithm must learn this perfect partition.
The algorithm's objective is to \emph{learn} the (static) communication graph
 while serving all requests,
and without executing too many migrations.
For the learning model, we  assume that the migration cost is $\alpha=1$.

@@

The two works closest to ours are by Avin et al. (on the general partitioning model) \cite{repartition-disc, sidma-arxiv} and by Henzinger et al. (on the learning model) \cite{sigmetrics19_partitioning}.
However, the focus of these papers is primarily on models with resource augmentation: the online algorithm can use slightly larger clusters than the offline algorithm.  
Avin et al. actually showed that their lower bound $\Omega(k)$ holds even for a significant resource augmentation, and they provided an algorithm with the competitive ratio $O(k \log k)$ using the $(2+\epsilon)$-augmented cluster capacity.
Their ratio is independent of $\ell$, which is impossible without significant resource augmentation.

In contrast, we study the non-augmented setting, where the nodes need to be perfectly balanced  among the clusters.
This assumption is not only more realistic but also significantly more challenging, as it is related to hard problems such as integer partitioning \cite{integer-partitions-book}.
In terms of results without augmentation, so far, it is only known that there exists a $O(k^2 \cdot \ell^2)$-competitive algorithm \cite{repartition-disc}; the best known lower bound is significantly lower, namely $\Omega(k)$.
For $k=2$, Avin et al.\cite{repartition-disc} presented a $7$-competitive algorithm with a~substantial ($\Omega(\ell^2)$) additive constant.

The problem has also been studied in a weaker
model where the adversary can only sample
requests from a fixed distribution~\cite{stochastic-ring}.

The static offline version of~the~partitioning~problem, i.e., a problem variant where
migration is not allowed, where all requests are known in advance, and where
the goal is to find an assignment of $n$ nodes to $\ell$~physical machines, each of~capacity $n/\ell$, is known as the
\emph{$\ell$-balanced graph partitioning problem}. The problem is 
NP-complete, and cannot even be approximated within any finite factor unless P
= NP~\cite{AndRae06}.  The static
variant where $\ell = 2$ corresponds to the minimum bisection problem, which
is already NP-hard~\cite{GaJoSt76}, and 
the currently best approximation ratio is $O(\log n)$~\cite{SarVaz95,ArKaKa99,FeKrNi00,FeiKra02,KraFei06,Raec08}.

Our problem is further related to some classic online problems.
In particular, it is related to online paging~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00}, sometimes also referred to
as online caching, where requests for data items (nodes) arrive over time and
need to be served from a cache of finite capacity, and where the number of
cache misses must be minimized. Classic problem variants usually boil down to
finding a smart eviction strategy, such as Least Recently Used (LRU)~\cite{SleTar85}. In our
setting, requests can be served remotely (i.e.,~without fetching the
corresponding nodes to a single physical machine). In this light, our model is more
reminiscent of caching models \emph{with
bypassing}~\cite{EpImLN11,EpImLN15,Irani02}.
A major difference between  these problems is that in the caching problems, each request involves a~single element of the universe, while in our model \emph{both} endpoints of a communication request are subject to~optimization.
In this light, we can see our model as a "symmetric" version of online paging.

Dynamic graph partitioning problems are generally fundamental in computer science, and arise in many different contexts \cite{streaming-soda,streaming1}.

%More generally, the model is related to online
%caching~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00},
%see~\cite{repartition-disc} for a discussion.
%The static offline version of~the~problem, called the
%\emph{$\ell$-balanced graph partitioning problem} is 
%NP-complete, and cannot even be approximated within %any finite factor unless P
%= NP~\cite{AndRae06}.

@@

This paper presents several new results on the dynamic graph partitioning problem  without augmentation.
For the learning model we present a lower bound of $\Omega(k\cdot\ell)$ on the competitive ratio of any online deterministic online algorithm 
(that holds also in the general partitioning model).
The best known lower bound so far was $\Omega(k)$~\cite{repartition-disc} that holds only in the general partitioning model.
We further present an asymptotically optimal, 
$O(k\cdot \ell)$-competitive algorithm
for the learning model.

For the general partitioning model, we design
an asymptotically optimal,
$\Theta(\ell)$-competitive algorithm for $k=3$, improving the best known upper bound 
so far $O(\ell^2)$~\cite{repartition-disc}.
We further present a strictly $6$-competitive algorithm for $k=2$ that improves upon previous $7$-competitive algorithm with $O(\alpha\ell^2)$ additive constant.

@@

\begin{table*}[h]
	\centering
	\renewcommand{\arraystretch}{1.5}
	\begin{tabular}{>{\centering\arraybackslash}p{4.25cm}|>{\centering\arraybackslash}p{4.25cm}>{\centering\arraybackslash}p{4.25cm}}
		\rowcolor{gray!50}
		\textbf{Variant} & \textbf{ Lower bound} &\textbf{Upper bound}\\ \hline 
		\textbf{$k=2$}& 3\hspace{0.3cm}\cite{repartition-disc} & 6\hspace{0.3cm}(\S \ref{sec:k2}) \\ 
		\rowcolor{gray!25}
		\textbf{$k=3$}&  $\Omega(\ell)$ \hspace{0.3cm}(\S \ref{sec:lowerbound})& $O(\ell) $\hspace{0.3cm}(\S \ref{sec:k3})\\
		$k > 3$ & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k^2 \cdot \ell^2)$\hspace{0.1cm} \cite{repartition-disc} \\
		\rowcolor{gray!25}
		Learning model & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k \cdot \ell)$\hspace{0.1cm} (\S \ref{sec:ppl}) \\
	\end{tabular}
	\caption{Overview of known results and our contributions. The table summarizes the results for the general partitioning model, except for the last row that summarizes the results for the learning model for arbitrary $k$ and $\ell$.
	}
	\label{tab:overview}
	\vspace{-7mm}
\end{table*}

\medskip
\noindent
\textbf{Algorithmic techniques.} The variant for general $k$ is still unresolved, hence it is vital to summarize algorithmic techniques used in this paper.
A straightforward analysis of the algorithm from Section~\ref{sec:k3} results in the bound $O(\ell^2)$, and the analysis must be improved in two places. One must bound the cost of each of $O(\ell)$ reconfigurations in a phase by a~constant, and show that the optimum offline algorithm must pay for a significant portion of inter-cluster requests \maciek{how to concisely say that there is also a migrate possibilty}.
We bound the cost of the latter by estimating the capabilities of the optimum offline algorithm to \emph{prepare} for an incoming sequence of requests.
Furthermore, in the analysis of the algorithm from Section~\ref{sec:k2}, we propose a~novel charging scheme for edges that share a vertex.

@@

In this section, we consider the learning variant of dynamic balanced graph partitioning problem.
For this setting, we show a surprisingly high lower bound of $\Omega(k \cdot \ell)$ for $k\geq 3$.
The lower bound holds also in the general partitioning model (studied in Section~\ref{sec:part}).
At the end of this section, we discuss an asymptotically optimal upper bound for the learning variant.

@@

We provide a lower bound $\Omega(k\cdot \ell)$ for the competitive ratio of any deterministic online algorithm for the learning problem.
Later we elaborate on how to efficiently transform it to a lower bound for the general partitioning problem.
The lower bound requires $k\geq 3$.
In contrast, for $k=2$ the learning problem is trivial: immediate collocation of communicating pairs is $1$-competitive.
(A partitioning problem for $k=2$ is non-trivial, a lower bound of $3$ is known~\cite{repartition-disc}, and we provide a~$6$-competitive algorithm, see Section~\ref{sec:k2}.)

Across this paper, we often refer to groups of communicating nodes.
We use this concept slightly differently in the lower bound and the upper bounds.
In our algorithms, we group nodes with a communication history into \emph{components}, and in the lower bound, we group nodes that may communicate into \emph{ground sets}.

A~\emph{ground set} is a group of nodes that repeatedly communicate when an algorithm splits them (places at different clusters).
Their role is to ensure that the algorithm maintains a~partition of ground sets into clusters.
We emphasize that ground sets are revealed only upon their nodes are split.
The adversary constructs ground sets depending on the choices of a deterministic online algorithm.
Once we construct a ground set, it lasts until the end of the algorithm's execution.
We say that a ground set is a \emph{singleton} if it contains exactly one node.
We say that the node that belongs to a singleton ground set is \emph{isolated}.

We start by constructing a ground set of size $k-1$ on an arbitrarily chosen cluster.
In any partition, there must exist an isolated node collocated with the ground set of size $k-1$, and we call it \emph{special}.
We issue requests between the current special node, and some node that was initially collocated with it.
Repeating such requests, almost all the nodes must visit the cluster $S$.
In comparison, we show that there exists an optimal offline algorithm \OPT that performs only two migrations.
Figure~\ref{fig:lb} illustrates the constructs used in the lower bound.

@@

\begin{proof}
	Fix any online algorithm \ALG{}.
	Let $I(C)$ denote the cluster where nodes of the ground set $C$ are located in the initial partition.
	If $C$ contains nodes that originate from different clusters, $I(C)$ is undefined.
	Throughout the construction, we maintain ground sets for each node.
	Initially, all nodes are isolated (i.e., in singleton ground sets).
	First, we create a ground set $B$ of $k-1$ nodes on an arbitrarily chosen cluster (the nodes  are already collocated).
	%We never dismiss any constructed ground set.
	Each cluster hosts exactly $k$ nodes, and in any partition, a single isolated node resides in $S$.
	We say that this isolated node that accompanies the ground set $B$ is \emph{special}.
	We denote the first special node by $x_0$.

	Then, we add the special node to a larger ground set to force its eviction.
	Precisely,
	we create a ground set $\{x_0, y_0\}$, 
	where $x_0$ is the current special node and $y_0$ is an arbitrary node that does not reside in $S$.
	\ALG has them split, and the adversary issues external requests until \ALG collocates them.
	There is no space to accommodate $\{x_0, y_0\}$ in the cluster of $x_0$ (as $B$'s size is $k-1$), hence \ALG must collocate $\{x_0, y_0\}$ in another cluster.
	To preserve a~feasible partition of nodes into clusters after collocating $\{x_0, y_0\}$,
	\ALG must replace $x_0$ with another single node that we call $x_1$.

	We continue adding the current special node to a larger ground set.
	Precisely, for an isolated node $x_i$ from~$S$, we merge the ground sets of $x_i$ and the largest ground set $C$ s.t. $C \neq \{x_0,y_0\}$ and $I(C) = I(x_i)$.
	This way, $x_i$ is evicted similarly to $x_0$, and we continue creating such ground sets.
	We describe the termination condition shortly.
	(Note that these ground sets contain nodes that originate from the same cluster, whereas the nodes $x_0$ and $y_0$ originate from distinct clusters.)

	We claim that expanding ground sets in this way can continue as long as at least $\ell$ isolated nodes exist.
	Fix the partition just before the creation of a ground set with $x_i$ for some~$i$.
	We show that if at least $\ell$ isolated nodes exist (including $x_i$), then there exists a~perfect partition of ground sets after expanding a ground set with $x_i$.
	To show this claim, we consider a partition, where the largest ground set of each cluster originates from it (excluding the ground set $\{x_0, y_0\}$).
	In this partition, each cluster contains:  at most one ground set larger than $1$, possibly some isolated nodes, and possibly the ground set $\{x_0, y_0\}$.
	If $I(x_i)$ contains an isolated node, then we simply obtain a~perfect partition by swapping it with $x_i$.
	In the following, we assume that no isolated nodes exist on $I(x_i)$.
	The largest ground set of $I(x_i)$ is at most $k-1$, because $x_i$ is on $S$.
	Therefore $\{x_0, y_0\}$ is present on $I(x_i)$, and we must evict it to make space for $x_i$.
	To find the cluster for $\{x_0, y_0\}$, observe that we have at most $\ell-1$ clusters with isolated nodes, and we have at least $\ell$ isolated nodes --- hence a~cluster with two isolated nodes exists.
	To find a feasible partition, we swap $\{x_0, y_0\}$ with these two collocated isolated nodes, and then we swap an isolated node from $I(x_i)$ with~$x_i$.

	To keep \OPT's cost low, we terminate prematurely --- we stop extending ground sets while there are still $\ell+1$ isolated nodes (i.e., neglecting two last ground set expansions).
	Consider \ALG's partition after the last expansion.
	Among $\ell+1$ isolated nodes there must exist a pair that originates from the same cluster (we have $\ell$ clusters). 
	We denote these two nodes by $x^*$ and $y^*$.

@@

This paper revisited the dynamic graph partitioning problem and presented several tight bounds for the important model where capacities cannot be exceeded, both for a general partitioning model and for a special learning model.

>>>>

\begin{abstract}
	Distributed   applications,  including  batch  processing, streaming, scale-out databases,
	or machine learning, generate a~significant amount of network traffic.
	By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), we can reduce the network load and  improve application performance. 
	However, the communication pattern of different applications is often unknown a~priori and may change over time, hence it needs to be learned in an online manner.
	%
	This paper revisits the online 
	balanced partitioning problem 
	(introduced by Avin et al.~at DISC 2016)
	that asks for an algorithm that strikes
	an optimal tradeoff between the benefits
	of collocation (i.e., lower network load) 
	and its costs (i.e., migrations). 
	%
	Our first contribution is a~significantly improved
	lower bound of $\Omega(k\cdot \ell)$ on the
	competitive ratio, where $\ell$ is the number
	of clusters and $k$ is the cluster size,
	even for a~scenario in which the communication
	pattern is static and can be perfectly partitioned;
	we also provide an asymptotically tight upper bound 
	of $O(k\cdot \ell)$ for this scenario.
	For $k=3$, we contribute an asymptotically tight upper bound
	of $\Theta(\ell)$
	for the general model in which the
	communication pattern can change arbitrarily over time.
	We improve the result for $k=2$ by providing a~strictly $6$-competitive upper bound for the general model.
	In contrast to most prior work, our algorithms respect all capacity constraints and do not require resource augmentation.

@@

The popularity of data-centric, distributed applications has led to an explosive growth of network traffic, especially in data centers~\cite{roy2015inside,singh2015jupiter}.
The performance of these distributed applications often critically depends on the underlying network~\cite{mogul2012we}, and efficient operation of these networks is important.
At the same time, distributed systems are often highly virtualized today, and provide interesting new opportunities for resource optimization.
In particular, it has become possible to operate data centers in a~more demand-aware manner: 
by dynamically migrating nodes (e.g., virtual machines) which communicate frequently topologically closer to each other, network traffic can be reduced significantly.  
However, migrations should be used moderately, as migrations also entail overheads.

This paper studies the algorithmic problem underlying such demand-aware
optimizations, aiming to strike a~balance between the benefits of migrations (e.g., reduced network load) and their costs.
In particular, we are interested in an online variant of the problem: since communication patterns can change over time, an online algorithm needs to react dynamically to new traffic patterns, and migrate nodes  accordingly.
Ideally, this algorithm should perform close to an optimal offline algorithm, without requiring any information about future traffic demands.

This problem is known as the dynamic balanced graph partitioning problem and was introduced by Avin et al.~\cite{repartition-disc, sidma-arxiv} at DISC 2016. a~special variant of the general problem has later been studied by Henzinger et al.~\cite{sigmetrics19_partitioning} at SIGMETRICS 2019.
We refer to the latter as the learning model.

@@

We study two models in this paper: the \emph{general partitioning} model, and its subproblem, the \emph{learning} model.
In both models, we assume that communication patterns are not known to our algorithms at the beginning.
We measure the~quality of~presented algorithmic solutions by competitive analysis~\cite{borodin-book}, which is well-suited for problems that are online by their nature.
In the competitive analysis, the goal is to~optimize \emph{the competitive ratio} of a~given online algorithm: the ratio of its cost to the cost of~an~optimal offline algorithm that knows the whole input sequence in advance.
%We emphasize that all our algorithms have a~\emph{strict} competitive ratio (without an additive term).

\noindent
\textbf{General partitioning model.}
In the \emph{dynamic balanced graph partitioning} problem, we are given a~set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a~sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a~pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

An algorithm serves a~communication request between two nodes
either \emph{locally} at cost~0
if they are collocated,
or \emph{remotely} at cost~1
if they are located in different clusters.
We refer to these two types of requests as \emph{internal}
and \emph{external} requests, respectively.
Before serving a~request,
an online algorithm may perform a~\emph{repartition},
%(i.e., \emph{reconfigure}).
i.e.,
it may move (``migrate'') some nodes into clusters different from their current clusters, while respecting the capacity of every cluster. 
Afterward, 
the algorithm serves the  request.
The cost of migrating a~node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $\ALG$,
its cost,
denoted by $\ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $\ALG$ while serving the sequence $\sigma$.

\noindent
\textbf{Learning model.}
We study a~\emph{learning} variant of dynamic balanced graph partitioning,
where the communication pattern is \emph{static}:
whether a~pair of  nodes ever communicate or not, 
is determined a~priori and is unknown to algorithms,
 and requests to such pairs arrive indefinitely.
Any algorithm must eventually collocate  pairs of communicating nodes,
as otherwise it cannot be competitive.
As in Henzinger et al.~\cite{sigmetrics19_partitioning}, we assume that the communication graph admits a~\emph{perfect partition},
i.e., a~partition in which no inter-cluster request ever occurs.
%any algorithm must learn this perfect partition.
The algorithm's objective is to \emph{learn} the (static) communication graph
 while serving all requests,
and without executing too many migrations.
For the learning model, we  assume that the migration cost is $\alpha=1$.

@@

The two works closest to ours are by Avin et al. (on the general partitioning model)~\cite{repartition-disc, sidma-arxiv} and by Henzinger et al. (on the learning model)~\cite{sigmetrics19_partitioning}.
However, the focus of these papers is primarily on models with resource augmentation: the online algorithm can use slightly larger clusters than the offline algorithm.  
Avin et al. actually showed that their lower bound $\Omega(k)$ holds even for a~significant resource augmentation, and they provided an algorithm with the competitive ratio $O(k \log k)$ using the $(2+\epsilon)$-augmented cluster capacity.
Their ratio is independent of $\ell$, which is impossible without significant resource augmentation.

In contrast, we study the non-augmented setting, where the nodes need to be perfectly balanced  among the clusters.
This assumption is not only more realistic but also significantly more challenging, as it is related to hard problems such as integer partitioning~\cite{integer-partitions-book}.
In terms of results without augmentation, so far, it is only known that there exists a~$O(k^2 \cdot \ell^2)$-competitive algorithm~\cite{repartition-disc}; the best known lower bound is significantly lower, namely $\Omega(k)$.
For $k=2$, Avin et al.~\cite{repartition-disc} presented a~$7$-competitive algorithm with a~substantial ($\Omega(\ell^2)$) additive constant.

The problem has also been studied in a~weaker
model where the adversary can only sample
requests from a~fixed distribution~\cite{stochastic-ring}.

The static offline version of~the~partitioning~problem, i.e., a~problem variant where
migration is not allowed, where all requests are known in advance, and where
the goal is to find an assignment of $n$ nodes to $\ell$~physical machines, each of~capacity $n/\ell$, is known as the
\emph{$\ell$-balanced graph partitioning problem}. The problem is 
NP-complete, and cannot even be approximated within any finite factor unless P
= NP~\cite{AndRae06}.  The static
variant where $\ell = 2$ corresponds to the minimum bisection problem, which
is already NP-hard~\cite{GaJoSt76}, and 
the currently best approximation ratio is $O(\log n)$~\cite{SarVaz95,ArKaKa99,FeKrNi00,FeiKra02,KraFei06,Raec08}.

Our problem is further related to some classic online problems.
In particular, it is related to online paging~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00}, sometimes also referred to
as online caching, where requests for data items (nodes) arrive over time and
need to be served from a~cache of finite capacity, and where the number of
cache misses must be minimized. Classic problem variants usually boil down to
finding a~smart eviction strategy, such as Least Recently Used (LRU)~\cite{SleTar85}. In our
setting, requests can be served remotely (i.e.,~without fetching the
corresponding nodes to a~single physical machine). In this light, our model is more
reminiscent of caching models \emph{with
bypassing}~\cite{EpImLN11,EpImLN15,Irani02}.
A major difference between  these problems is that in the caching problems, each request involves a~single element of the universe, while in our model \emph{both} endpoints of a~communication request are subject to~optimization.
In this light, we can see our model as a~"symmetric" version of online paging.

Dynamic graph partitioning problems are generally fundamental in computer science, and arise in many different contexts~\cite{streaming-soda,streaming1}.

%More generally, the model is related to online
%caching~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00},
%see~\cite{repartition-disc} for a~discussion.
%The static offline version of~the~problem, called the
%\emph{$\ell$-balanced graph partitioning problem} is 
%NP-complete, and cannot even be approximated within %any finite factor unless P
%= NP~\cite{AndRae06}.

@@

This paper presents several new results on the dynamic graph partitioning problem  without augmentation.
For the learning model we present a~lower bound of $\Omega(k\cdot\ell)$ on the competitive ratio of any online deterministic online algorithm 
(that holds also in the general partitioning model).
The best known lower bound so far was $\Omega(k)$~\cite{repartition-disc} that holds only in the general partitioning model.
We further present an asymptotically optimal, 
$O(k\cdot \ell)$-competitive algorithm
for the learning model.

For the general partitioning model, we design
an asymptotically optimal,
$\Theta(\ell)$-competitive algorithm for $k=3$, improving the best known upper bound 
so far $O(\ell^2)$~\cite{repartition-disc}.
We further present a~strictly $6$-competitive algorithm for $k=2$ that improves upon previous $7$-competitive algorithm with $O(\alpha\ell^2)$ additive constant.

@@

\begin{table*}[h]
	\centering
	\renewcommand{\arraystretch}{1.5}
	\begin{tabular}{>{\centering\arraybackslash}p{4.25cm}|>{\centering\arraybackslash}p{4.25cm}>{\centering\arraybackslash}p{4.25cm}}
		\rowcolor{gray!50}
		\textbf{Variant} & \textbf{ Lower bound} &\textbf{Upper bound}\\ \hline 
		\textbf{$k=2$}& 3\hspace{0.3cm}\cite{repartition-disc} & 6\hspace{0.3cm}(\S \ref{sec:k2}) \\ 
		\rowcolor{gray!25}
		\textbf{$k=3$}&  $\Omega(\ell)$ \hspace{0.3cm}(\S \ref{sec:lowerbound})& $O(\ell) $\hspace{0.3cm}(\S \ref{sec:k3})\\
		$k > 3$ & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k^2 \cdot \ell^2)$\hspace{0.1cm}~\cite{repartition-disc} \\
		\rowcolor{gray!25}
		Learning model & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k \cdot \ell)$\hspace{0.1cm} (\S \ref{sec:ppl}) \\
	\end{tabular}
	\caption{Overview of known results and our contributions. The table summarizes the results for the general partitioning model, except for the last row that summarizes the results for the learning model for arbitrary $k$ and $\ell$.
	}
	\label{tab:overview}
	\vspace{-7mm}
\end{table*}

\medskip
\noindent
\textbf{Algorithmic techniques.} The variant for general $k$ is still unresolved, hence it is vital to summarize algorithmic techniques used in this paper.
A straightforward analysis of the algorithm from Section~\ref{sec:k3} results in the bound $O(\ell^2)$, and the analysis must be improved in two places. One must bound the cost of each of $O(\ell)$ reconfigurations in~phase by a~constant, and show that the optimum offline algorithm must pay for a~significant portion of inter-cluster requests \maciek{how to concisely say that there is also a~migrate possibilty}.
We bound the cost of the latter by estimating the capabilities of the optimum offline algorithm to \emph{prepare} for an incoming sequence of requests.
Furthermore, in the analysis of the algorithm from Section~\ref{sec:k2}, we propose a~novel charging scheme for edges that share a~vertex.

@@

In this section, we consider the learning variant of dynamic balanced graph partitioning problem.
For this setting, we show a~surprisingly high lower bound of $\Omega(k \cdot \ell)$ for $k\geq 3$.
The lower bound holds also in the general partitioning model (studied in Section~\ref{sec:part}).
At the end of this section, we discuss an asymptotically optimal upper bound for the learning variant.

@@

We provide a~lower bound $\Omega(k\cdot \ell)$ for the competitive ratio of any deterministic online algorithm for the learning problem.
Later we elaborate on how to efficiently transform it to a~lower bound for the general partitioning problem.
The lower bound requires $k\geq 3$.
In contrast, for $k=2$ the learning problem is trivial: immediate collocation of communicating pairs is $1$-competitive.
(A partitioning problem for $k=2$ is non-trivial, a~lower bound of $3$ is known~\cite{repartition-disc}, and we provide a~$6$-competitive algorithm, see Section~\ref{sec:k2}.)

Across this paper, we often refer to groups of communicating nodes.
We use this concept slightly differently in the lower bound and the upper bounds.
In our algorithms, we group nodes with a~communication history into \emph{components}, and in the lower bound, we group nodes that may communicate into \emph{ground sets}.

A~\emph{ground set} is a~group of nodes that repeatedly communicate when an algorithm splits them (places at different clusters).
Their role is to ensure that the algorithm maintains a~partition of ground sets into clusters.
We emphasize that ground sets are revealed only upon their nodes are split.
The adversary constructs ground sets depending on the choices of a~deterministic online algorithm.
Once we construct a~ground set, it lasts until the end of the algorithm's execution.
We say that a~ground set is a~\emph{singleton} if it contains exactly one node.
We say that the node that belongs to a~singleton ground set is \emph{isolated}.

We start by constructing a~ground set of size $k-1$ on an arbitrarily chosen cluster.
In any partition, there must exist an isolated node collocated with the ground set of size $k-1$, and we call it \emph{special}.
We issue requests between the current special node, and some node that was initially collocated with it.
Repeating such requests, almost all the nodes must visit the cluster $S$.
In comparison, we show that there exists an optimal offline algorithm \OPT that performs only two migrations.
Figure~\ref{fig:lb} illustrates the constructs used in the lower bound.

@@

\begin{proof}
	Fix any online algorithm \ALG{}.
	Let $I(C)$ denote the cluster where nodes of the ground set $C$ are located in the initial partition.
	If $C$ contains nodes that originate from different clusters, $I(C)$ is undefined.
	Throughout the construction, we maintain ground sets for each node.
	Initially, all nodes are isolated (i.e., in singleton ground sets).
	First, we create a~ground set $B$ of $k-1$ nodes on an arbitrarily chosen cluster (the nodes  are already collocated).
	%We never dismiss any constructed ground set.
	Each cluster hosts exactly $k$ nodes, and in any partition, a~single isolated node resides in $S$.
	We say that this isolated node that accompanies the ground set $B$ is \emph{special}.
	We denote the first special node by $x_0$.

	Then, we add the special node to a~larger ground set to force its eviction.
	Precisely,
	we create a~ground set $\{x_0, y_0\}$, 
	where $x_0$ is the current special node and $y_0$ is an arbitrary node that does not reside in $S$.
	\ALG has them split, and the adversary issues external requests until \ALG collocates them.
	There is no space to accommodate $\{x_0, y_0\}$ in the cluster of $x_0$ (as $B$'s size is $k-1$), hence \ALG must collocate $\{x_0, y_0\}$ in another cluster.
	To preserve a~feasible partition of nodes into clusters after collocating $\{x_0, y_0\}$,
	\ALG must replace $x_0$ with another single node that we call $x_1$.

	We continue adding the current special node to a~larger ground set.
	Precisely, for an isolated node $x_i$ from~$S$, we merge the ground sets of $x_i$ and the largest ground set $C$ s.t. $C \neq \{x_0,y_0\}$ and $I(C) = I(x_i)$.
	This way, $x_i$ is evicted similarly to $x_0$, and we continue creating such ground sets.
	We describe the termination condition shortly.
	(Note that these ground sets contain nodes that originate from the same cluster, whereas the nodes $x_0$ and $y_0$ originate from distinct clusters.)

	We claim that expanding ground sets in this way can continue as long as at least $\ell$ isolated nodes exist.
	Fix the partition just before the creation of a~ground set with $x_i$ for some~$i$.
	We show that if at least $\ell$ isolated nodes exist (including $x_i$), then there exists a~perfect partition of ground sets after expanding a~ground set with $x_i$.
	To show this claim, we consider a~partition, where the largest ground set of each cluster originates from it (excluding the ground set $\{x_0, y_0\}$).
	In this partition, each cluster contains:  at most one ground set larger than $1$, possibly some isolated nodes, and possibly the ground set $\{x_0, y_0\}$.
	If $I(x_i)$ contains an isolated node, then we simply obtain a~perfect partition by swapping it with $x_i$.
	In the following, we assume that no isolated nodes exist on $I(x_i)$.
	The largest ground set of $I(x_i)$ is at most $k-1$, because $x_i$ is on $S$.
	Therefore $\{x_0, y_0\}$ is present on $I(x_i)$, and we must evict it to make space for $x_i$.
	To find the cluster for $\{x_0, y_0\}$, observe that we have at most $\ell-1$ clusters with isolated nodes, and we have at least $\ell$ isolated nodes --- hence a~cluster with two isolated nodes exists.
	To find a~feasible partition, we swap $\{x_0, y_0\}$ with these two collocated isolated nodes, and then we swap an isolated node from $I(x_i)$ with~$x_i$.

	To keep \OPT's cost low, we terminate prematurely --- we stop extending ground sets while there are still $\ell+1$ isolated nodes (i.e., neglecting two last ground set expansions).
	Consider \ALG's partition after the last expansion.
	Among $\ell+1$ isolated nodes there must exist a~pair that originates from the same cluster (we have $\ell$ clusters). 
	We denote these two nodes by $x^*$ and $y^*$.

@@

This paper revisited the dynamic graph partitioning problem and presented several tight bounds for the important model where capacities cannot be exceeded, both for a~general partitioning model and for a~special learning model.
